---
title: Template
---

## Markdown

We assume a base knowledge of Markdown.

The knitting process uses Pandoc. Please refer to the [Pandoc documtentation](https://pandoc.org/MANUAL.html#pandocs-markdown) for supported Markdown syntax.

## YAML

YAML is a serialized language comprised of key value pairs, in essence, this means spaces are significant. Key value pairs are seperated by a colon and space `: ` and indenting of additional arguments uses a tab. The entire thing is wrapped in three back dashes `---`. Now that we're using Quarto, things should be relatively simple, and all you need is a title.

```markdown
---
title: page title
---
```

## Quarto

Code cells in Quarto are generally created in the same way as in RMarkdown, i.e.

```{r}
#| echo: fenced

dat <- palmerpenguins::penguins
```

Code cell options are set globally in `_quarto.yml`. Currently `warning` and `error` are set to `false`.

Additional cell options are set at the top of the chunk with `#| option: argument`. These accept tab auto completion.

```{r}
#| echo: fenced
#| warning: true

library(dplyr)
```

Additional execution options can be found at [https://quarto.org/docs/computations/execution-options.html](https://quarto.org/docs/computations/execution-options.html)

## Callout Blocks

Callout blocks are created with the following syntax:

```markdown
::: {callout-option}
Your text here.
:::
```

The callout options that we will be using include:

| Use case | Option |
| :--- | :--- |
| Your Turn | .callout-caution |
| Warning | .callout-warning | 
| Walk Along | .callout-tip |
| Note | .callout-note |

An example:

```markdown
::: {.callout-tip}
Your note here.
:::
```

Will render as:

::: {.callout-tip}
Your note here.
:::

Note the title of the callout box is not what we want though. We can't change this globally. The easiest way to address is this is to use a level two header inside the callout.

An example:

```markdown
::: {.callout-tip}
## Walk Along

Your walk along here.
:::
```

Will render as:

::: {.callout-tip}
## Walk Along

Your walk along here.
:::
